---
title: "PA1_template.Rmd"
author: "Allison Harney"
date: "June 22, 2018"
output: html_document
---
#Reproducible Research: Project 1
```{r setup, include=TRUE}
knitr::opts_chunk$set(echo = TRUE)

library(ggplot2)
library(dplyr)
```

##**Loading and preprocessing the data**

###1. Code for reading in the dataset and/or processing data

```{r loading}
if(!file.exists('activity.csv')){
    unzip('activity.zip')
}
activity <- read.csv('activity.csv')

```
##**What is mean total number of steps taken per day?**
###2. Histogram of the total number of steps taken each day
```{r histogram}
steps <- activity %>%
        group_by(date)%>%
        summarize(stepsperday = sum(steps))
ggplot(steps, aes(stepsperday)) + geom_histogram(bins = 10) + 
    xlab("Total steps") + ylab("Frequency") + ggtitle("Total Number of Steps Per Day")

```

###3.Mean and median number of steps taken each day 
Mean number of steps per day with NA removed
```{r mean}
stepsByDayMean <- mean(steps$stepsperday, na.rm = TRUE)
print(stepsByDayMean)

```
Median number of steps per day with NA removed
```{r median}
stepsByDayMedian <- median(steps$stepsperday, na.rm = TRUE)
print(stepsByDayMedian)

```
The mean number of steps taken each day was `r stepsByDayMean`.
The median number of steps taken each day was `r stepsByDayMedian`.

##**What is the average daily pattern?**
###4. Time series plot of the average number of steps taken
```{r time_series}
time <- activity %>%
        group_by(interval)%>%
        summarize(meanstepsperday = mean(steps, na.rm = TRUE))
ggplot(time, aes(y = meanstepsperday, x = interval)) + geom_line() + 
    xlab("Interval") + ylab("Mean Number of Steps") + ggtitle("Average Daily Activity Pattern")

```

###5. The 5-minute interval that, on average, contains the maximum number of steps
```{r max}
five <- time %>%
    filter(meanstepsperday == max(meanstepsperday))
print(five)

```
The 5-minute interval of `r five$interval` on average contains the maximum number of steps per days with `r five$meanstepsperday` steps per day.

##**Imputing missing values**
###6. Code to describe and show a strategy for imputing missing data
```{r missing}
missing <- activity %>%
    summarize(sum(is.na(activity)))
print(missing)

```
There are `r missing` missing values in the data set. 
For all missing values, the average number of steps from the entire data set will be used. 
```{r imputed}
imputed <- activity %>%
    mutate(steps = replace(steps, is.na(steps), mean(steps, na.rm = TRUE)))

```

###7. Histogram of the total number of steps taken each day after missing values are imputed
```{r imputed_histo}
steps_imputed <- imputed %>%
            group_by(date)%>%
            summarize(stepsperday = sum(steps))
ggplot(steps_imputed, aes(stepsperday)) + geom_histogram(bins = 10) + 
    xlab("Total steps") + ylab("Frequency") + ggtitle("Total Number of Steps Per Day")

meanImputed <- mean(steps_imputed$stepsperday)
print(meanImputed)

medianImputed <- median(steps_imputed$stepsperday)
print(medianImputed)
```
The mean number of steps per day in the imputed data set is `r meanImputed`.
The median number of steps per day in the imputed data set is `r medianImputed`.

The mean number of steps per day remains the same after imputation as the original dataset. The median is slightly higher such that the median now equals the mean value. 

##**Are there differences in activity pattern between weekdays and weekends?**

###8. Panel plot comparing the average number of steps taken per 5-minute interval across weekdays and weekends
```{r imputed_days}
imputed <- mutate(imputed, date = as.Date(date))

weekdays <- imputed%>%
    mutate(day = weekdays(date))%>%
    mutate(weekend = ifelse(day == "Saturday" | day == "Sunday", "weekend", "weekday") )

weekdays_steps <- weekdays %>%
            group_by(weekend, interval)%>%
            summarize(meanstepsperday = mean(steps, na.rm = TRUE))

ggplot(weekdays_steps, aes(y = meanstepsperday, x = interval))  + geom_line() + facet_wrap(~weekend) +
    xlab("Interval") + ylab("Mean Number of Steps") 


```
